 

### 2021-02-01 vertical-thermal-review.Rmd
Minor edits to draft above line 260

Rising temperatures are expected to have profound effects on forests; however, it is not well understood how responses will vary across forest strata.
Here, we synthesize evidence as to how environmental conditions and foliar traits vary across vertical gradients, shaping leaf temperatures, metabolism, and ultimately whole-tree demography and ecosystem function.
Strong vertical microclimatic gradients imply that canopy leaves are exposed to more solar radiation and higher evaporative demand than understory leaves, particularly in forests with dense canopies.
However, foliar traits relevant to shaping leaf temperature ($T_{leaf}$) and metabolism also vary strongly across height or light gradients.
Resulting $T_{leaf}$'s exceed air temperature ($T_{air}$) under conditions of high solar radiation, low wind, and low stomatal conductance.
Differences are most pronounced when hydraulic limitations impede evaporative cooling such that $T_{leaf}$'s of sun-exposed canopy leaves become dramatically elevated above both $T_{air}$ and understory $T_{leaf}$.
While leaf metabolism generally increases with height across the vertical gradient, differences in thermal sensitivity, optimal temperatures, and thermal damage thresholds are modest.
The implications for tree performance are mixed: whereas thermal buffering may contribute to the disproportionate stress of larger trees during drought, emerging tree-ring evidence suggests that understory trees experience relatively greater stress from high $T_{air}$.
Additional research will be important for improving our spotty understanding of the thermal sensitivity of metabolism and tree growth across vertical forest gradients.
In the meantime, our findings imply that while large canopy trees are the most vulnerable to warming when combined with drought, understory trees may be more vulnerable under more mesic conditions.

@@

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**
Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. 
Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory. 
This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack synthetic understanding of these gradients. 
*(@niinemetsWithincanopyVariationRate2004a highlights how biophysical and photosynthetic gradients vary along the vertical canopy profile, however our current knowledge of this topic is lacking.)*
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

**Here we review how the biophysical environment and biological form and function vary across the vertical canopy gradient in forests.** 
We focus on five key themes (Fig. `r fig_schematic`):
(1) the biophysical environment;
(2) the leaf traits that most strongly influence leaf temperature and metabolism;
(3) resulting leaf temperature ($T_{leaf}$) and
(4) foliar metabolism, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
(5) resultant size-structuring of whole-tree and ecosystem function.
We then consider the implications for modeling /  climate change responses.

@@

### Light environments
**Light conditions, specifically changes in proportion incident light and photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig. `r fig_NEON_vertical`; @koike_leaf_2001].** 
Generally, upper canopies in dense forests are exposed to greater PAR values and limit light penetration to canopy interiors, lower canopy layers, and the understory. Shaded regions experience lower values of PAR with different light quality [in ranges of blue, orange and red]. The light gradient is more pronounced under broadleaf forests than needleleaf forests [@lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @niinemetsPackingPhotosyntheticMachinery2009; @poorterMetaanalysisPlantResponses2019b]
(Fig. `r fig_NEON_vertical`).

In heterogeneous canopies [with high gap fractions], the distance from the outer canopy [to X?] is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gaps to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019]

Tropical forests with densely packed overstory trees limit light to subsequent canopy layers and forest understories. Tropical forest floors receive approximately 1-2%  of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006]. However, vertical canopy layers and forest floors also receive light permeating through canopy gaps, as well as intermittent light penetration from tree, branch, and leaf movement caused by wind, defined collectively as sunflecks. Sunflecks can provide 10-80% of photon flux density for photosynthesis in canopy layers and understory, depending on canopy attributes such as foliage density, height and architecture [@waySunflecksTreesForests2012]

@@

RH, in combination with temperature, determines vapor pressure deficit, $VPD$. Atlantic forests in Brazil during the wet season tend to have higher $T_{air}$, and $VPD$ in the upper canopy and canopy gaps than in the understory where $T_{air}$ and $VPD$ tend to be lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].  
This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomatal constraints [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]. In northeastern US mesic temperate forests, tree-ring analyis [__time-scale] showed that the upper canopy was observed to be more sensitive to $VPD$, whereas the understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]

@@

**Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [@rozendaalPlasticityLeafTraits2006].**
After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody and herbaceous species [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a]. A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR, and the ratio of red to far red radiation [R:FR] intercepting the leaf surface. This enables acclimation or avoidance responses through the development of necessary leaf traits [@casalPhotoreceptorSignalingNetworks2013]. Degrees of high irradiance give rise to variation in sun-leaf traits, while degrees of low irradiance-- characterized by shade-- result in variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; @sackHowStrongIntracanopy2006], ultimately influencing leaf structural, chemical and physiological traits as well [@keenanGlobalLeafTrait2016].

@@

Light based acclimation in leaf traits is well established, while research on solely height-based leaf trait variation is sparse due to the difficulty of isolating increased tree height from light. However, studies that have measured LMA in relation to height have found that in tropical forests LMA-height predicted stronger correlation between area- based respiration versus foliar nutrients, whereas Q10 and E10 remained constant with height [Cavaleri et al. 2008]. In temperate Acer saccharum forest, within-canopy Acer saccharum exhibited greater LMA and $N area$ with height early in the growing season as a result of hydrostatic constraints on leaf morphology; however, in later growing season LMA and $N area$ continued to increase in the upper canopy due to light driven changes [@cobleLightDrivesVertical2014a; @cobleLightAcclimationOptimizes2015; @cobleHowVerticalPatterns2016].In addition, carbon isotope composition also increased with height, independent of light [@cobleLightAcclimationOptimizes2015]

>>>>

Rising temperatures are expected to have profound effects on forests; however, it is not well understood how responses will vary across forest strata.
Here, we synthesize evidence as to how environmental conditions and foliar traits vary across vertical gradients, shaping leaf temperatures, metabolism, and ultimately whole-tree growth and mortality.
Strong vertical microclimatic gradients imply that canopy leaves are exposed to more solar radiation and higher evaporative demand than understory leaves, particularly in forests with dense canopies.
However, foliar traits relevant to shaping leaf temperature ($T_{leaf}$) and metabolism also vary strongly across height or light gradients.
Resulting $T_{leaf}$'s exceed air temperature ($T_{air}$) under conditions of high solar radiation, low wind, and low stomatal conductance.
Differences are most pronounced when hydraulic limitations impede evaporative cooling such that $T_{leaf}$'s of sun-exposed canopy leaves become dramatically elevated above both $T_{air}$ and understory $T_{leaf}$.
While leaf metabolism generally increases with height across the vertical gradient, differences in thermal sensitivity, optimal temperatures, and thermal damage thresholds are modest.
The implications for tree performance are mixed: whereas thermal buffering may contribute to the disproportionate stress of larger trees during drought, emerging tree-ring evidence suggests that understory trees experience relatively greater stress from high $T_{air}$.
Additional research will be important for improving our spotty understanding of the thermal sensitivity of metabolism and tree growth across vertical forest gradients.
In the meantime, our findings imply that while large canopy trees are the most vulnerable to warming when combined with drought, understory trees may be more vulnerable under more mesic conditions.

@@

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**
Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. 
Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory. 
This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack synthetic [replace synthetic w comprehensive?] understanding of these gradients. 
*(@niinemetsWithincanopyVariationRate2004a highlights how biophysical and photosynthetic gradients vary along the vertical canopy profile, however our current knowledge of this topic is lacking.)*
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

**Here, we review biophysical and biological patterns across the vertical canopy gradient in forests, how these gradients affect leaf-level processes and, in turn, whole plant performance , and the implications for understanding forest ecosystem responses and climate change feedbacks.**
We structure the review around five major questions (Fig. `r fig_schematic`).

i.	How does the biophysical environment vary with height in forests?

ii.	How do the leaf traits that most strongly influence leaf temperature and metabolism vary with height (or between sun and shade leaves) in forests?

iii. How do vertical changes in biophysical environment conditions and leaf traits interact to affect leaf temperature [and in turn photosynthetic capacity]??

iv.	How does leaf metabolism respond to [air or leaf?] temperature in canopy and understory settings?

v. What are the implications of vertical canopy gradient patterns for the ecological and climate change responses of canopy versus understory trees?

We then consider the implications for ecosystem-scale dynamics and climate change feedbacks, as well as remote sensing and modeling opportunities for future research.

@@

### Light environments
**Light conditions, specifically changes in the proportion of incident light and photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig. `r fig_NEON_vertical`; @koike_leaf_2001].** 
Generally, upper canopies in dense forests are exposed to greater PAR values and limit light penetration to canopy interiors, lower canopy layers, and the understory. Shaded regions experience lower values of PAR with different light quality [in ranges of blue, orange and red]. The light gradient is more pronounced under broadleaf forests than needleleaf forests [@lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @niinemetsPackingPhotosyntheticMachinery2009; @poorterMetaanalysisPlantResponses2019b]
(Fig. `r fig_NEON_vertical`).

In heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy [to X?] is a better proxy for light environment than height (Parker 1995). Lower canopy light environments are highly variable, ranging from high light gaps to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019]

Tropical forests with densely packed overstory trees limit light reaching subsequent canopy layers and forest understories. Tropical forest floors receive approximately 1-2%  of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006]. However, vertical canopy layers and forest floors also receive light permeating through canopy gaps, as well as intermittent light penetration from tree, branch, and leaf movement caused by wind, defined collectively as sunflecks. Sunflecks can provide 10-80% of photon flux density for photosynthesis in canopy layers and understory, depending on canopy attributes such as foliage density, height and architecture [@waySunflecksTreesForests2012]

@@

RH, in combination with temperature, determines vapor pressure deficit ($VPD$), an important metric expressing the driving force of water loss from a leaf. Atlantic forests in Brazil during the wet season tend to have higher $T_{air}$, and $VPD$ in the upper canopy and canopy gaps than in the understory where $T_{air}$ and $VPD$ tend to be lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017]. This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomatal constraints [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]. In northeastern US mesic temperate forests, tree-ring analyis [__time-scale] showed that the upper canopy was observed to be more sensitive to $VPD$, whereas the understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]

@@

**Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [@rozendaalPlasticityLeafTraits2006].**
After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody and herbaceous species [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a]. A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR and the ratio of red to far red radiation [R:FR] intercepting the leaf surface. This enables acclimation or avoidance responses through the development of necessary leaf traits [@casalPhotoreceptorSignalingNetworks2013]. Degrees of high irradiance give rise to variation in sun-leaf traits, while degrees of low irradiance-- characterized by shade-- result in variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; @sackHowStrongIntracanopy2006], ultimately influencing leaf structural, chemical and physiological traits as well [@keenanGlobalLeafTrait2016].

@@

Light based acclimation in leaf traits is well established, while research on solely height-based leaf trait variation is sparse due to the difficulty of isolating increased tree height from light. However, studies that have measured LMA in relation to height have found that in tropical forests LMA-height predicted stronger correlation between area- based respiration versus foliar nutrients, whereas Q10 and E10 remained constant with height [Cavaleri et al. 2008]. In a temperate Acer saccharum forest, within-canopy Acer saccharum exhibited greater LMA and $N area$ with height early in the growing season as a result of hydrostatic constraints on leaf morphology; however, in later growing season LMA and $N area$ continued to increase in the upper canopy due to light driven changes [@cobleLightDrivesVertical2014a; @cobleLightAcclimationOptimizes2015; @cobleHowVerticalPatterns2016].In addition, carbon isotope composition also increased with height, independent of light [@cobleLightAcclimationOptimizes2015]
